Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-11-26 manuscript.tex
rerun pipeline

When treating head and neck squamous cell carcinomas (HNSCC) with
radiotherapy or surgery, not only the primary tumor and the clinically
detected lymph node metastases are targeted. Per current guidelines,
large volumes of the neck are part of the elective clinical target
volume (CTV-N)
\citep{gregoire_ct-based_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensity-modulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}.
This is to minimize the risk of regional recurrences due to untreated
microscopic disease which cannot be detected by current in-vivo imaging
modalities such as computed tomography (CT), magnetic resonance imaging
(MRI), or positron emission tomography (PET). However, minimizing the
risk of missing occult disease in the lymph drainage region must be
balanced against the toxicity related to unnecessary treatment of
healthy tissue.

The aforementioned guidelines for the definition of the CTV-N are based
on anatomically defined lymph node levels (LNL)
\citep{gregoire_delineation_2014} and the overall prevalence of lymph
node metastases in the LNLs. Based on this, they recommend extensive
irradiation of both sides of the neck for a majority of patients.
However, the overall prevalence of involvement in a given LNL does not
equal a patient's personal risk for occult disease in that LNL, which
depends on an individual patient's state of tumor progression.
Currently, a patient presenting with no clinically detectable nodal
disease (cN0) and a small, clearly lateralized T1 tumor, the same
recommended contralateral CTV-N as a patient with substantial nodal
involvement in the ipsilateral side and an advanced primary tumor that
crosses the mid-sagittal plane. Both patients receive extensive elective
irradiation of the contralateral LNLs II, III, and IVa
\citep{biau_selection_2019}.

To better quantify this personalized risk of occult disease we have
previously developed an intuitive, probabilistic hidden Markov model
\citep{ludwig_hidden_2021, ludwig_modelling_2023} which was originally
based on a conceptually similar Bayesian network model
\citep{pouymayou_bayesian_2019}. So far, these models were limited to
the ipsilateral side of the neck. The main contribution of this work is
the extension of the formalism to include risk predictions for the
contralateral side as well. The model may then guide the reduction of
the electively irradiated volumes in the contralateral neck for patients
whose individual risk for occult disease is low. In turn, this may
substantially reduce the toxicity from irradiation and the associated
decrease in quality of life.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  In section section~\ref{sec-data} we present a multi-centric dataset
  on lymph node involvement in 833 OPSCC patients. Based on the dataset
  the main risk factors for contralateral lymph node involvement are
  identified and the requirements for the bilateral model extension are
  described (section~\ref{sec-requirements}).
\item
  In section~\ref{sec-ext-to-contra} we present a bilateral HMM of
  lymphatic progression that accounts for the T-stage lateralization of
  the primary tumor as well as clinical involvement as risk factors for
  occult contralateral involvement. In section~\ref{sec-methods} we
  describe how the model training and the computational experiments were
  set up.
\item
  In the results in section~\ref{sec-results}, we demonstrate that the
  model is able to describe the patterns of contralateral lymph node
  involvement observed in the dataset, and we apply the model to
  estimating risk of occult disease for typical patients. In
  section~\ref{sec-discussion}, implications for volume-deescalated
  radiotherapy are discussed.
\end{enumerate}

@@

To be able to create models for lymphatic tumor progression that include
all relevant LNLs including the contralateral side, we have collected a
detailed dataset of 833 patients with newly diagnosed oropharyngeal
squamous cell carcinomas
\citep{ludwig_dataset_2022, ludwig_multi-centric_2023}. It reports the
lymph node involvement per LNL of every patient individually in tabular
form, in addition to other primary tumor and patient characteristics
such as T-category, subsite, lateralization of the primary tumor, and
HPV p16 status. Their patient records have been collected at four
different institutions and a brief overview over some of their patients'
characteristics are shown in table~\ref{tbl-data-overview}. The data
from the Inselspital Bern (ISB) and the Centre Léon Bérard (CLB) only
consist of patients who all received a neck dissection. The majority of
patients from the University Hospital Zürich (USZ) and the Hospital Vall
d'Hebron (HVH) was treated with definitive radiotherapy. Since sugical
treatment is more common for early T-category patients, the ISB and CLB
datasets contain a larger portion of early T-category patients compared
to the USZ and HVH dataset. For a subset of 83 patients in the CLB
dataset, the lateralization of the primary tumor was not reported.

@@

\caption{}\label{T_d2cbf}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

Pathological involvement is available only for the subset of surgical
patients and only for the dissected levels. For the remaining patients,
only clinical involvement is available. For the analysis performed in
this paper, all available diagnostic information has been combined into
a consensus decision representing the most likely state of involvement
for each patient and LNL, using literature values for sensitivity and
specificity of different modalities
\citep{de_bondt_detection_2007, kyzas_18f-fluorodeoxyglucose_2008}. How
this consensus is formed is described in detail in
section~\ref{sec-consensus}. In brief, pathology after neck dissection
is taken as the ground truth if available, overwriting any contradicting
clinical involvement diagnoses. If a level was not dissected, the most
likely state is determined by PET-CT in most cases.

@@

The entire data, including additional patients with tumors in other
primary locations than the oropharynx, is publicly available: It may be
\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
it can be interactively explored too,
\href{https://github.com/rmnldwg/lydata}{from GitHub},
\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
\emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
\citet{ludwig_multi-centric_2023}. The Data-in-Brief publications
contain a detailed description of the datasets and the data format.
These publications do not include the most recent dataset addition from
HVH, which will be described in a future publication.

@@

These datasets allow us to investigate correlations between the
involvement of individual LNLs, or between risk factors and involvement.
In figure~\ref{fig-data-strat}, we have plotted the prevalence of each
contralateral LNL's involvement, stratified by the risk factors
T-category, number of ipsilaterally involved LNLs, and whether the tumor
extended over the mid-sagittal plane.

@@

The bottom left panel in figure~\ref{fig-data-strat} shows that patients
with a tumor crossing the mid-sagittal plane show contralateral
involvement vastly more often compared to patients with clearly
lateralized tumors. This makes intuitive sense, because the lymphatic
system in the head and neck region is typically symmetric and no major
lymph vessels cross the midline. Therefore, interstitial fluids from the
primary tumor -- which we assume to carry living malignant cells -- may
only reach the blind-ended lymphatic vessels in the contralateral neck
via short-ranged diffusion. Which in turn is only possible when the
primary tumor is close enough to the mid-sagittal line or crosses it.

The top left panel in figure~\ref{fig-data-strat} indicates that
T-category is correlated with contralateral involvement (as it is with
overal involvement). This is because T-category may be considered a
surrogate for the time between onset of disease and diagnosis. I.e., a
patient with a T4 tumor was -- on average -- diagnosed later than a
patient with a T1 tumor. Thus, the former did have more time to develop
metastases.

Similarly, ipsilateral involvement correlates with contralateral
metastasis (top right panel). The tumor of a patient with many
metastases in ipsilateral LNLs was probably able to spread for longer
(or faster) compared to a tumor in a patient with no nodal disease. The
degree of ipsilateral involvement, too, may therefore be considered a
surrogate for the duration of the disease. In addition, it has been
hypothesized that bulky nodal disease ipsilaterally may also redirect
lymph fluids to the contralateral side.

The three risk factors for contralateral involvement, midline extension,
T-category, and ipsilateral involvement are correlated. For example,
midline extension was present in 45.56\% of advanced T-category tumors
but only in 6.13\% of early T-category tumors. The higher prevalence of
contralateral metastases in advanced T-category tumors may thus be
partially explained by a higher rate of midline extension in these
tumors. However, the bottom right panel of figure~\ref{fig-data-strat}
indicates that advanced T-category and severe ipsilateral involvement do
represent additional risk factors. The figure considers only patients
with lateralized tumors not extending over the midline. For early
T-category and healthy ipsilateral levels I-V, only 1.85\% (2 out of 108
patients) shows involvement of contralateral level II. This increases to
6.73\% (14 out of 208 patients) if ipsilateral level II is involved, to
14.29\% (10 out of 70 patients) if ipsilateral levels II and III are
involved, and to 39.68\% (25 out of 63 patients) for advanced T-category
with ipsilateral levels II and III involved.

@@

Based on the observations of the \hyperref[sec-data-strat]{previous
section}, any model to predict the risk for contralateral nodal
involvement, should be able to describe the following:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A tumor that extends over the mid-sagittal line should yield
  contralateral metastases with much higher probability.
\item
  More advanced T-category should lead to higher risk for nodal disease.
  One approach to achieve this via the expected time of diagnosis has
  already been developed in the form of a hidden Markov model
  \citep{ludwig_hidden_2021}.
\item
  The degree of ipsilateral involvement should give the model
  information on the time that may have passed between onset and
  diagnosis of the disease. This should come in addition to what can be
  inferred about the time from T-category alone.
\end{enumerate}

@@

This paper builds on the previously developed unilateral model for
ipsilateral lymph node involvement presented in {[}cite 2024 graph
extension paper{]}. In this section we provide a brief recap of the
unilateral model to introduce the notation needed to extend the
framework to a bilateral model describing both ipsilateral and
contralateral lymph node involvement in section~\ref{sec-ext-to-contra}.
For further details in the ipsilateral model, the reader is referred to
the earlier publications \citep[cite 2024 graph extension
paper]{ludwig_hidden_2021}

We model a patient's state of involvement at an abstract time-step \(t\)
as a vector of hidden binary random variables:

@@

Here, \(V\) is the number of LNLs the model considers. The values a
LNL's hidden binary random variables may take on are \(X_v[t] = 0\)
(\texttt{False}), meaning the LNL \(v\) is healthy or free of metastatic
disease, or \(X_v[t] = 1\) (\texttt{True}), corresponding to some form
of tumor presence (i.e., occult or clinically detected). Since the state
vector \(\mathbf{X}[t]\) is \(V\)-dimensional and binary, there are
\(2^V\) distinct possible lymphatic involvement patterns, which we
enumerate from
\(\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}\)
to
\(\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}\).
In addition, each LNL is associated with an observed binary random
variable \(Z_v\) that describes the clinical involvement of a LNL based
on imaging: \(Z_v = 0\) (\texttt{False}) indicates that the LNL \(v\) is
healthy based on clinical diagnosis, and \(Z_v = 1\) (\texttt{True})
indicates that suspicious lymph nodes were detected that are deemed
metastatic.

Any hidden Markov model is fully described by three quantities:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A starting state \(\mathbf{X}[t=0]\) at time \(t=0\) just before the
  patient's tumor formed. In our case, this is always the state where
  all LNLs are still healthy \(\boldsymbol{\xi}_0\).
\item
  The \emph{transition matrix}
  \begin{equation}\phantomsection\label{eq-trans-matrix}{
  \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
  }\end{equation} where the value at row \(i\) and column \(j\)
  represents the probability to transition from state
  \(\boldsymbol{\xi}_i\) to \(\boldsymbol{\xi}_j\) during the time-step
  from \(t\) to \(t+1\). Note that we prohibit self-healing, meaning
  that during a transition, no LNL may change their state from
  \(X_v[t]=1\) to \(X_v[t+1]=0\). Consequently, many elements of the
  transition matrix are zero.
\item
  Lastly, the \emph{observation matrix}
  \begin{equation}\phantomsection\label{eq-obs-matrix}{
  \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
  }\end{equation} where in row \(i\) and at column \(j\) we find the
  probability to \emph{observe} a lymphatic involvement pattern
  \(\mathbf{Z} = \boldsymbol{\zeta}_j\), given that the true (but
  hidden) state of involvement at the time of diagnosis \(t_D\) is
  \(\mathbf{X}[t_D] = \boldsymbol{\xi}_i\).
\end{enumerate}

The transition matrix \(\mathbf{A}\) is parametrized using a directed
acyclic graph (DAG) as an abstract representation of the underlying
lymphatic network. Directed arcs from the primary tumor to a LNL are
associated with a probability \(b_v\) for direct spread to LNL \(v\)
during one time step. Directed arcs from a LNL \(v\) to a LNL \(r\) are
associated with a probability \(t_{vr}\) for the tumor to progress from
one LNL to the next. In this paper, we build on the DAG shown in
figure~\ref{fig-full-graph} which was obtained by maximizing the model
evidence as described in {[}cite 2024 graph extension paper{]}.

@@

Using the introduced quantities, we can evolve the distribution of all
possible hidden states from \(\mathbf{X}[t=0] = \boldsymbol{\xi}_0\)
step by step, by successively multiplying this vector with the
transition matrix \(\mathbf{A}\). For later use, we define at this point
the matrices of the dsitributions over all hidden states, given all
time-steps \(\boldsymbol{\Lambda}\):

@@

At the time of the diagnosis \(t_D\), we multiply the result with the
observation matrix \(\mathbf{B}\). We may then look up the likelihood of
a patient presenting with the diagnosis
\(\mathbf{Z}=\boldsymbol{\zeta}_i\) in the \(i\)-th entry of the final
result. However, the remaining issue is that the value of \(t_D\) is
unknown, i.e.~over how many time-steps the HMM should be evolved. We
solve this problem by marginalizing over the time of diagnosis.
Different distributions over the diagnosis times can then be choosen
based on T-category. For instance, the mean of the time-prior to
marginalize over the diagnosis time for early T-category patients
\(P\left( t_D \mid \text{early} \right)\) may be shifted towards earlier
times than the one for advanced T-category patients
\(P\left( t_D \mid \text{early} \right)\). This gives us for example

@@

With the formalism introduced above, we can write the likelihood
function for a patient to present with a diagnosis consisting of an
observed state and a T-category
\(d = \left( \boldsymbol{\zeta}_i, \text{T}x \right)\) as follows:

@@

Above, the quantity inside \(\left[ \ldots \right]_i\) denotes the
\(i\)-th component of the vector that is the result of the vector and
matrix multiplications in the square brackets. Note that it is also
possible to account for missing involvement information: If a diagnosis
(like fine needle aspiration (FNA)) is only available for a subset of
all LNLs, we can sum over all those possible complete observed states
\(\boldsymbol{\zeta}_j\) that match the provided diagnosis.

@@

And it is our goal to infer the values of these parameters for a given
dataset \(\mathcal{D} = \left( d_1, d_2, \ldots, d_N \right)\) of OPSCC
patients. The likelihood of these \(N\) diagnoses is simply the product
of their individual likelihoods as defined in
equation~\ref{eq-single-patient-llh}. For numerical reasons, we
typically compute the data likelihood in log space:

@@

A naive approach to model the contralateral lymphatic spread would be to
simply employ two independent unilateral models as introduced in
section~\ref{sec-unilateral}. During training, one could enforce that
some parameters are shared between these two models, e.g.~the
parameterization of the distributions over diagnose times or the spread
among the LNLs (\$t\_\{vr\}). However, this approach lacks a way to
describe the correlation between ipsi- and contralateral involvement
discussed in section \{\#sec-data-strat\}. Two independent unilateral
models would not describe the observation that contralateral involvement
becomes more likely with more severe ipsilateral involvement.

Thus, we extend the formalism in section~\ref{sec-unilateral} in such a
way that the model's ipsi- and contralateral side evolve synchronously
over time. To achieve that, we start by writing down the posterior
distribution of involvement an analogy to
equation~\ref{eq-uni-bayes-law}, which is now a joint probability of an
involvement \(\mathbf{X}^\text{i}\) ipsilaterally \emph{and} an
involvement \(\mathbf{X}^\text{c}\) contralaterally, given a diagnosis
of the ipsilateral LNLs \(\mathbf{Z}^\text{i}\) and of the contralateral
ones \(\mathbf{Z}^\text{c}\):

@@

The term representing the model's prior probability of hidden
involvement does not factorize. However, we assume that there is no
direct lymph drainage from an ipsilateral LNL to a contralateral LNL. No
major lymph vessels cross the mid-sagittal plane. In the graphical
model, this means that we assume no directed arcs between ipsilateral
and contralateral LNLs, i.e.~tumor spread to the contralateral side is
due to the primary tumor alone. We can thus write the joint probability
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)\) as a
factorising sum:

@@

In general, the matrices \(\boldsymbol{\Lambda}_\text{i}\) and
\(\boldsymbol{\Lambda}_\text{c}\) could be parameterized using a
disjoint set of parameters, i.e., the ipsi- and contralateral spread
rates are entirely different. However, using three sensible assumptions,
we can reduce the parameter space by sharing some parameters between the
two sides:

>>>>

In treating head and neck squamous cell carcinomas (HNSCC) with
radiotherapy or surgery, both the primary tumor and clinically detected
lymph node metastases are targeted. Current guidelines also include
large portions of the neck in the elective clinical target volume
(CTV-N)
\citep{gregoire_ctbased_2003, gregoire_delineation_2014, gregoire_delineation_2018, eisbruch_intensitymodulated_2002, biau_selection_2019, chao_determination_2002, vorwerk_guidelines_2011, ferlito_elective_2009}
to mitigate the risk of regional recurrences from untreated microscopic
disease undetectable by in-vivo imaging modalities such as computed
tomography (CT), magnetic resonance imaging (MRI), or positron emission
tomography (PET). However, this approach must balance minimizing the
risk of occult disease in the lymphatic drainage region against the
toxicity of unnecessarily irradiating healthy tissue.

These CTV-N guidelines rely on anatomically defined lymph node levels
(LNLs) \citep{gregoire_delineation_2014} and the overall prevalence of
lymph node metastases within these levels. They often recommend
extensive irradiation of both sides of the neck. However, the general
prevalence of metastasis in a given LNL does not correspond to an
individual patient's risk of occult disease in that region, which
depends on their specific state of tumor progression. For example, a
patient with no clinically detectable nodal disease (cN0) who has a
small, clearly lateralized T1 tumor would receive the same contralateral
CTV-N as a patient with significant ipsilateral nodal involvement and an
advanced tumor crossing the mid-sagittal plane: Both patients receive
elective irradiation of the contralateral LNLs II, III, and IVa
\citep{biau_selection_2019}.

To better quantify individualized risk of occult disease, we previously
developed an intuitive probabilistic hidden Markov model (HMM)
\citep{ludwig_hidden_2021, ludwig_modelling_2023}, originally based on a
conceptually similar a Bayesian network model
\citep{pouymayou_bayesian_2019}. However, these models have been limited
to predicting ipsilateral nodal involvement. This work extends the model
to include contralateral risk predictions, enabling more personalized
radiation volume recommendations. By identifying patients with low
contralateral risk, the model could guide reductions in contralateral
CTV-N, thereby decreasing radiation-induced toxicity and improving
quality of life.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Section section~\ref{sec-data} presents a multi-centric dataset on
  lymph node involvement in 833 OPSCC patients, identifying key risk
  factors for contralateral lymph node involvement and outlining
  requirements for a bilateral model extension
  (section~\ref{sec-requirements}).
\item
  Section section~\ref{sec-ext-to-contra} introduces a bilateral HMM
  that incorporates primary tumor lateralization, T-category, and
  clinical involvement as risk factors for contralateral involvement.
  Model training and computational experiments are described in
  section~\ref{sec-methods}.
\item
  Section section~\ref{sec-results} demonstrates the model's ability to
  replicate observed contralateral lymph node involvement patterns and
  estimates occult disease risk for typical patients. Implications for
  volume-deescalated radiotherapy are discussed in
  section~\ref{sec-discussion}.
\end{enumerate}

@@

To develop models for lymphatic tumor progression for all relevant LNLs,
including contralateral regions, we compiled a detailed dataset of 833
patients with newly diagnosed oropharyngeal squamous cell carcinomas
(OPSCC) \citep{ludwig_dataset_2022, ludwig_multicentric_2024}. The
dataset includes lymph node involvement per LNL for each patient in
tabular form, along with primary tumor and patient characteristics such
as T-category, subsite, primary tumor lateralization, and HPV p16
status. Patient records were collected from four institutions, and an
overview of patient characteristics is provided in
table~\ref{tbl-data-overview}.

Data from Inselspital Bern (ISB) and Centre Léon Bérard (CLB) consist
exclusively of patients who underwent neck dissections. In contrast, the
majority of patients from the University Hospital Zürich (USZ) and the
Hospital Vall d'Hebron (HVH) were treated with definitive radiotherapy.
Since surgical treatment is more common for early T-category patients,
ISB and CLB datasets include a higher proportion of these cases compared
to USZ and HVH. For 83 patients in the CLB dataset, the primary tumor's
lateralization was not reported.

@@

\caption{}\label{T_561f9}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

Pathological involvement is available only for surgically treated
patients and for the levels that were dissected. For non-surgical
patients, involvement status is determined clinically, i.e.~using
imaging. To ensure uniformity in analysis, diagnostic information was
synthesized into a consensus decision for each patient and LNL. This
consensus reflects the most likely state of involvement and accounts for
the sensitivity and specificity of various diagnostic modalities, as
reported in the literature
\citep{debondt_detection_2007, kyzas_18ffluorodeoxyglucose_2008}.

The consensus process is detailed in section~\ref{sec-consensus}.
Briefly, pathological findings from neck dissections are treated as the
gold standard, overriding any conflicting clinical diagnoses. For levels
not dissected, PET-CT is typically the primary source for determining
the most likely state of involvement.

@@

The complete dataset, including additional patients with tumors in
primary sites other than the oropharynx, is publicly accessible. It can
be \href{https://lyprox.org/patients/dataset}{downloaded from LyProX},
where it is also available for interactive exploration,
\href{https://github.com/rmnldwg/lydata}{from GitHub}, or
\href{https://zenodo.org/search?q=lydata}{from Zenodo}. Additional
details about the datasets and data format are provided in designated
Data-in-Brief publications
\citep{ludwig_dataset_2022, ludwig_multicentric_2024}. These
publications do not include the latest data from HVH, which will be
described in a future publication.

@@

The datasets enable analysis of correlations between LNL involvement and
key risk factors. In figure~\ref{fig-data-strat}, we illustrate the
prevalence of contralateral LNL involvement, stratified by T-category,
the number of ipsilaterally involved LNLs, and whether the tumor extends
across the mid-sagittal plane.

@@

\subsubsection{Midline Extension}\label{midline-extension}

The bottom left panel of figure~\ref{fig-data-strat} shows that tumors
crossing the mid-sagittal plane have a substantially higher prevalence
of contralateral involvement compared to clearly lateralized tumors.
This aligns with the anatomy of the head and neck lymphatic system,
which is symmetric, with no major lymph vessels crossing the midline.
Interstitial fluids from the primary tumor, presumed to carry malignant
cells, can merely diffuse to contralateral lymphatic vessels over short
distances. Thus, contralateral spread is more likely when the tumor
approaches or crosses the midline.

\subsubsection{T-Category}\label{t-category}

The top left panel shows a correlation between T-category and
contralateral involvement, reflecting T-category's role as a surrogate
for the time elapsed between disease onset and diagnosis. Advanced
T-category tumors (e.g., T4) generally represent a longer disease
progression timeline, providing more opportunity for metastatic spread
compared to smaller tumors (e.g., T1).

\subsubsection{Ipsilateral Involvement}\label{ipsilateral-involvement}

The top right panel reveals a positive correlation between ipsilateral
and contralateral metastases. Extensive ipsilateral involvement likely
indicates a longer or faster disease progression. Additionally, it has
been hypothesized that bulky ipsilateral nodal disease may reroute
lymphatic drainage toward the contralateral side, potentially
facilitating contralateral metastasis.

\subsubsection{Interplay of Risk
Factors}\label{interplay-of-risk-factors}

Midline extension, T-category, and ipsilateral involvement are
interrelated risk factors for contralateral metastasis. For instance,
45.56\% of advanced T-category tumors exhibit midline extension compared
to 6.13\% of early T-category tumors. While the higher fraction of
midline extensions in advanced T-category patients partially accounts
for the higher contralateral metastasis rates, T-category itself and
ipsilateral involvement also play an additional role.

The bottom right panel of figure~\ref{fig-data-strat} isolates patients
with lateralized tumors that do not cross the midline, further
demonstrating these relationships. Among early T-category patients with
no ipsilateral nodal involvement (levels I-V), only 1.85\% (2 of 108
patients) show involvement in contralateral level II. This proportion
increases to 6.73\% (14 of 208) if ipsilateral level II is involved, to
14.29\% (10 of 70) if ipsilateral levels II and III are involved, and
further to 39.68\% (25 of 63) for advanced T-category tumors with
ipsilateral levels II and III involved.

@@

Based on the observations in section~\ref{sec-data-strat} above, any
model predicting the risk of contralateral nodal involvement should
account for the following:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Midline Extension:} Tumors extending across the mid-sagittal
  plane should result in a significantly higher probability of
  contralateral metastases.
\item
  \textbf{T-Category and Disease Progression:} Advanced T-category
  should correspond to an increased risk of nodal disease. This can be
  modeled using the expected time of diagnosis, as demonstrated in a
  prior hidden Markov model \citep{ludwig_hidden_2021}.
\item
  \textbf{Ipsilateral Involvement as a Progression Marker:} The extent
  of ipsilateral involvement should provide additional information on
  the time elapsed between disease onset and diagnosis, supplementing
  the insights gained from T-category alone.
\end{enumerate}

@@

This paper builds on the previously developed unilateral model for
ipsilateral lymph node involvement presented in
\citep{ludwig_modelling_2024}. Below, we briefly recap the unilateral
model to introduce the notation required for extending the framework to
a bilateral model, described in section~\ref{sec-ext-to-contra}. For
further details on the ipsilateral model, refer to earlier publications
\citep{ludwig_hidden_2021, ludwig_modelling_2024}.

We represent a patient's state of involvement at an abstract time-step
\(t\) as a vector of hidden binary random variables, where each
component corresponds to a lymph node level (LNL):

@@

Here, \(V\) is the number of LNLs the model considers. The values a
LNL's may take on are \(X_v[t] = 0\) (\texttt{False}), meaning the LNL
\(v\) is healthy or free of metastatic disease, or \(X_v[t] = 1\)
(\texttt{True}), corresponding to some form of tumor presence (i.e.,
occult or macroscopic disease). In total, there are \(2^V\) distinct
possible lymphatic involvement patterns, which we enumerate from
\(\boldsymbol{\xi}_0 = \begin{pmatrix} 0 & 0 & \cdots & 0 \end{pmatrix}\)
to
\(\boldsymbol{\xi}_{2^V} = \begin{pmatrix} 1 & 1 & \cdots & 1 \end{pmatrix}\).
Each LNL's state is observed via another binary random variable \(Z_v\)
that describes the clinical involvement of a LNL based on imaging:
\(Z_v = 0\) (\texttt{False}) indicates that the LNL \(v\) is healthy
based on clinical diagnosis, and \(Z_v = 1\) (\texttt{True}) indicates
that LNL \(v\) was classified as involved. \(X_v\) and \(Z_v\) are
connected through the sensitivity and specificity of the diagnositc
modality.

Based on this, our HMM is fully described by defining the following
three quantities:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A starting state \(\mathbf{X}[t=0]\) at time \(t=0\) just before the
  patient's tumor formed. In our case, this is always the state
  \(\boldsymbol{\xi}_0\) where all LNLs are still healthy.
\item
  The \emph{transition matrix}
  \begin{equation}\phantomsection\label{eq-trans-matrix}{
  \mathbf{A} = \left( A_{ij} \right) = \big( P \left( \mathbf{X}[t+1] = \boldsymbol{\xi}_j \mid \mathbf{X}[t] = \boldsymbol{\xi}_i \right) \big)
  }\end{equation} where the value at row \(i\) and column \(j\)
  represents the probability to transition from state
  \(\boldsymbol{\xi}_i\) to \(\boldsymbol{\xi}_j\) during the time-step
  from \(t\) to \(t+1\). Note that we prohibit self-healing, meaning
  that during a transition, no LNL may change their state from
  \(X_v[t]=1\) to \(X_v[t+1]=0\). Consequently, many elements of the
  transition matrix are zero.
\item
  Lastly, the \emph{observation matrix}
  \begin{equation}\phantomsection\label{eq-obs-matrix}{
  \mathbf{B} = \left( B_{ij} \right) = \big( P \left( \mathbf{Z} = \boldsymbol{\zeta}_j \mid \mathbf{X}[t_D] = \boldsymbol{\xi}_i \right) \big)
  }\end{equation} where in row \(i\) and at column \(j\) we find the
  probability to \emph{observe} a lymphatic involvement pattern
  \(\mathbf{Z} = \boldsymbol{\zeta}_j\), given that the true (but
  hidden) state of involvement at the time of diagnosis \(t_D\) is
  \(\mathbf{X}[t_D] = \boldsymbol{\xi}_i\).
\end{enumerate}

The transition matrix \(\mathbf{A}\) is parameterized using a directed
acyclic graph (DAG) that represents the underlying lymphatic network.
Edges from the primary tumor to an LNL are associated with a probability
\(b_v\) for direct spread to LNL \(v\) during one time step. Arcs from a
LNL \(v\) to a LNL \(r\) are parameterized with the probability rate
\(t_{vr}\) for inter-nodal progression. In this paper, we build on the
DAG shown in figure~\ref{fig-full-graph} which was obtained by
maximizing the model evidence as described in
\citep{ludwig_modelling_2024}.

@@

We can evolve the distribution of all possible hidden states from
\(\mathbf{X}[t=0] = \boldsymbol{\xi}_0\) step by step, by successively
multiplying this vector with the transition matrix \(\mathbf{A}\):
\(\mathbf{X}[t+1] = \mathbf{X}[t] \cdot \mathbf{A}\). For later use, we
define at this point a matrix \(\boldsymbol{\Lambda}\) that collects
these distributions for all considered time-steps:

@@

At the time of diagnosis, \(0 \leq t_D \leq t_\text{max}\), we multiply
the evolved state distribution with the observation matrix
\(\mathbf{B}\) to obtain the distribution over all possible diagnoses.
However, the exact time of diagnosis, \(t_D\), is unknown; that is, we
do not know the number of time-steps over which the HMM should be
evolved. To address this, we marginalize over all possible diagnosis
times, allowing the diagnosis to occur at any time-step, albeit with
different weights. These weights are defined by a prior distribution
over \(t_D\), which can vary depending on the patient's T-category. For
example, the time-prior for early T-category patients,
\(P(t_D \mid \text{early})\), may put more weight on earlier time-steps,
reflecting -- on average -- earlier detection, compared to the prior for
advanced T-category patients, \(P(t_D \mid \text{advanced})\).

@@

The likelihood function for a patient to present with a diagnosis
\(\boldsymbol{\zeta}_i\) a T-category \(\text{T}x\) tumor can now be
written as:

@@

With \(\left[ \ldots \right]_i\) we denote the \(i\)-th component of the
vector in the square brackets. Note that it is also possible to account
for missing involvement information: If a diagnosis (like fine needle
aspiration (FNA)) is only available for a subset of all LNLs, we can sum
over all those possible complete observed states
\(\boldsymbol{\zeta}_j\) that match the provided diagnosis.

@@

And it is our goal to infer optimal parameter values of from a given
dataset \(\mathcal{D}\) (consisting of diagnoses and T-categories) of
OPSCC patients. The likelihood to observe this cohort of \(N\) patients,
given a set of parameters \(\boldsymbol{\theta}\) is simply the product
of their individual likelihoods as defined in
equation~\ref{eq-single-patient-llh}. For numerical reasons, we
typically compute the data likelihood in log space:

@@

A straightforward approach to modeling contralateral lymphatic spread
would be to use two independent unilateral models, as described in
Section section~\ref{sec-unilateral}, with shared parameters such as the
distribution of diagnosis times or spread between LNLs (\(t_{vr}\)).
However, this method would fail to capture the correlation between
ipsilateral and contralateral involvement discussed in
section~\ref{sec-data-strat}, particularly the observed increase in
contralateral involvement with greater severity of ipsilateral spread.

Thus, we extend the formalism in section~\ref{sec-unilateral} in such a
way that the model's ipsi- and contralateral side evolve synchronously
over time. To achieve that, we start by writing down the posterior
distribution of involvement in analogy to
equation~\ref{eq-uni-bayes-law}, which is now a joint probability of an
involvement \(\mathbf{X}^\text{i}\) ipsilaterally \emph{and} an
involvement \(\mathbf{X}^\text{c}\) contralaterally, given a diagnosis
of the ipsilateral LNLs \(\mathbf{Z}^\text{i}\) and of the contralateral
ones \(\mathbf{Z}^\text{c}\):

@@

The term representing the model's prior probability of hidden
involvement does not factorize. However, we assume no direct lymphatic
drainage from ipsilateral to contralateral LNLs, as major lymph vessels
do not cross the mid-sagittal plane. In the graphical model, this
translates to the absence of directed arcs between ipsilateral and
contralateral LNLs, implying that contralateral tumor spread occurs
solely via the primary tumor. We can thus write the joint probability
\(P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)\) as a
factorising sum:

@@

The matrices \(\boldsymbol{\Lambda}\text{i}\) and
\(\boldsymbol{\Lambda}\text{c}\) could, in principle, be parameterized
with entirely separate parameters, allowing ipsilateral and
contralateral spread rates to differ completely. However, we simplify
the parameter space by sharing parameters between the two sides, based
on the following three assumptions:
